A Computational Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-04-02 paper/paper.tex
improve beginning

\section{Introduction}
%
When left unperturbed, the human brain is not at rest.
Rather, the brain continues to metabolize large quantities of
oxygen and glucose energy to maintain
potentially purposeful neuronal computation
in the absence of a focused behavioral goal
\citep{kenet2003spontaneously, fiser2004small}.
This baseline energy demand is subject to surprisingly little modulations
when mentally processing environmental challenges
\citep{raichle2001pnas}.
What has early been described as the "stream of consciousness"
in psychology \citep{james1890principles}
found its neurobiological manifestation
in the so-called "default mode network" (DMN)
\citep{shul1997}. This networked collection of some of the highest regions in
the human association cortex \citep{mesulam1998sensation, margulies2016situating}
consistently
increased in neural activity during
unfocused everyday mind wandering \citep{raichle2001pnas}.
Indeed, the human brain features "a hierarchy of brain systems with
the DMN at the top and the salience and dorsal attention systems
at intermediate levels, above thalamic and unimodal sensory
cortex" \citep{carhart2010default}.
In the beginning of the 21st century, brain imaging was hence the first technology
to allow for the discovery of a unique brain network that subserves
baseline mental activities
\citep{bzdok2015resting}.

\subsection{Default mode network: Higher-order control of the organism}
The DMN appeared to be exclusive in task-induced
deactivation during various psychological experiments.
This processing of unknown information categories in the DMN has
been argued to mediate an evolutionarily conserved function for the individual.
This is all the more likely because the DMN contains the two
biggest hotspots of energy consumption in the entire central nervous system
\citep{raichle2001pnas}.
DMN activity also persists to a substantial degree during
sleep and under anesthesia \citep{randy2008}.
Today, many authors believe that the DMN implements
probabilistic estimation of past, hypothetical, and future events. This
brain network
might have emerged to continuously predict environmental events using
mental imagery as an evolutionary advantage.
%
However, information processing in the DMN has also repeatedly
been shown to directly impact human behavior. Goal-directed task performance
improved with decreased activity in default mode areas \citep{weiss2006}
and increased DMN activity was linked to more task-independent,
yet sometimes useful thoughts
\citep{mason2007, seli2016mind}.
%
Understanding of the overarching DMN function is complicated by
simultaneously controlling perception-action cycles and
maintaining baseline contemplations
across time, splace, and content domains
at the interface of the external world and the self.

Consider an agent faced with the choice of the next action
and being guided by rich
reenactment of really-happened, hypothetically conceived, and
predicted future events to perfect ongoing behavioral performance.
A particularly attractive computational framework
to describe, quantify, and predict autonomously acting systems like the brain
can be the combination of control theory and reinforcement learning.
It is known that, the more the external world is predictable,
the more mental activity becomes detached from the actual sensory environment
\citep{antrobus1966studies, pope1978regulation}.
Conversely,
the more the currently executed task is unknown and unpracticed,
the lower the tendency for stimulus-independent thoughts
\citep{filler1973daydreaming, teasdale1995stimulus}.
These ``offline'' processes may however contribute to optimized control of the organism.
% We formalize
% a policy matrix to capture the space of possible actions that can be performed
% on the environment given the current state. A value matrix
% maps expected rewards to environmental objects and events.
% Switching between states reduces to a sequential processing model.
% In a model-based reinforcement learning (RL) approach,
% we try to understand the environment always better
% starting, at birth, with an environment that is completely unknown.
Informed by outcomes of performed action,
the DMN dynamics are constantly adapted in feedback loops
that are calibrated by prediction error.

A DMN framework incorporating RL can naturally embed human behavior
into the tension between exploitative action with immediate gains and
exploratory action with longer-term reward schedules.

\subsection{Sketch of our main contributions}
Our theory proposes that the implication of the DMN in a
diversity of humans' most advanced cognitive processes
can be parsimoniously recast as prediction error minimization
based on pervasive probabilistic simulations,
thus maximizing reward outcome at various time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Markov Chain Monte Carlo sampling
\citep{tenenbaum2011grow}.
\textit{Control} refers to the influences that an agent exerts when interacting
with the environment to encourage preferred states.
Even (necessarily imperfect) memories
of past experience, random mind-wandering and dreams during sleep
may provide meaningful building blocks to iteratively improve
the predictive DMN machinery to optimize the behavior of the organism.
%
In short, it has been proposed earlier that
the human brain's energy budget is largely dedicated to
``the development and maintenance of [a]
probabilistic model of anticipated events''
\citep{raichle2005intrinsic}.
The idea has been invigorated by empirical evidence from
neuroscientific experiments \citep{kording2004bayesian, fiser2004small}.
This paper proposes a formal model that satisfies this contention.

>>>>

\section{Introduction}
%
When left unperturbed, the human brain is not at rest.
Rather, the brain continues to metabolize large quantities of
oxygen and glucose energy to maintain
potentially purposeful neuronal computation
in the absence of a focused behavioral goal
\citep{kenet2003spontaneously, fiser2004small}.
This baseline energy demand is subject to surprisingly little modulations
when mentally processing environmental challenges
\citep{raichle2001pnas}.
What has early been described as the "stream of consciousness"
in psychology \citep{james1890principles}
found a potential neurobiological manifestation
in the so-called "default mode network" (DMN)
\citep{shul1997}. This networked collection of some of the highest regions in
the human association cortex \citep{mesulam1998sensation, margulies2016situating}
consistently
increased in neural activity during
unfocused everyday mind wandering \citep{raichle2001pnas}.
The DMN may exert higher-order control on behavior
as the human brain was argued to features "a hierarchy of brain systems with
the DMN at the top and the salience and dorsal attention systems
at intermediate levels, above thalamic and unimodal sensory
cortex" \citep{carhart2010default}.
In the beginning of the 21st century,
brain imaging may have been first technique
to allow for the discovery of a unique brain network that subserves
baseline mental activities
\citep{raichle2001pnas, bzdok2015resting}.

\subsection{Default mode network: Higher-order control of the organism}
The DMN appeared to be exclusive in task-induced
deactivation during various psychological experiments.
This processing of unknown information categories in the DMN has
been argued to mediate an evolutionarily conserved function for the individual.
This is all the more likely because the DMN contains hotspots
of highest energy consumption in the entire central nervous system
\citep{raichle2001pnas}.
DMN activity also persists to a substantial degree during
sleep and under anesthesia \citep{randy2008}.
Today, many authors believe that the DMN implements
probabilistic estimation of past, hypothetical, and future events. This
brain network
might have emerged to continuously predict environmental events using
mental imagery as an evolutionary advantage.
%
However, information processing in the DMN has also repeatedly
been shown to directly impact human behavior. Goal-directed task performance
improved with decreased activity in default mode areas \citep{weiss2006}
and increased DMN activity was linked to more task-independent,
yet sometimes useful thoughts
\citep{mason2007, seli2016mind}.
%
Understanding of the overarching DMN function is complicated by
simultaneously controlling perception-action cycles and
maintaining baseline contemplations
across time, splace, and content domains
at the interface of the external world and the self.

Consider an agent faced with the choice of the next action
and being guided by rich
reenactment of really-happened, hypothetically conceived, and
predicted future events to perfect ongoing behavioral performance.
A particularly attractive computational framework
to describe, quantify, and predict autonomously acting systems like the brain
can be the combination of control theory and reinforcement learning.
It is known that, the more the external world is predictable,
the more mental activity becomes detached from the actual sensory environment
\citep{antrobus1966studies, pope1978regulation}.
Conversely,
the more the currently executed task is unknown and unpracticed,
the lower the tendency for stimulus-independent thoughts
\citep{filler1973daydreaming, teasdale1995stimulus}.
These ``offline'' processes may however contribute to optimized control of the organism.
% We formalize
% a policy matrix to capture the space of possible actions that can be performed
% on the environment given the current state. A value matrix
% maps expected rewards to environmental objects and events.
% Switching between states reduces to a sequential processing model.
% In a model-based reinforcement learning (RL) approach,
% we try to understand the environment always better
% starting, at birth, with an environment that is completely unknown.
Informed by outcomes of performed action,
the DMN dynamics are constantly adapted in feedback loops
that are calibrated by prediction error.
A DMN framework incorporating RL can naturally embed human behavior
into the tension between exploitative action with immediate gains and
exploratory action with longer-term reward schedules.

\subsection{Sketch of our main contributions}
We argue that the implication of the DMN in a
diversity of humans' most advanced cognitive processes
can be parsimoniously recast as prediction error minimization
based on pervasive probabilistic simulations,
thus maximizing reward outcome at various time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Markov Chain Monte Carlo sampling
\citep{tenenbaum2011grow}.
\textit{Control} refers to the influences that an agent exerts when interacting
with the environment to encourage preferred states.
Even (necessarily imperfect) memories
of past experience, random mind-wandering and dreams during sleep
may provide meaningful building blocks to iteratively improve
the predictive DMN machinery to optimize the behavior of the organism.
%
In short, it has been proposed earlier that
the human brain's energy budget is largely dedicated to
``the development and maintenance of [a]
probabilistic model of anticipated events''
\citep{raichle2005intrinsic}.
The idea has been invigorated by empirical evidence from
neuroscientific experiments \citep{kording2004bayesian, fiser2004small}.
This paper proposes a formal model that satisfies this contention.
